Inferring the effective start dates of non-pharmaceutical interventions during COVID-19 outbreaks

### 2020-05-19 ms/ms.tex
fix fig 4

However, a significant difference between the beginning of the intervention and the effective change in transmission rates can invalidate assessments that assume a serial interval of 4-7 days and neglect the late or early population response to the NPI.
Such a case is illustrated in \autoref{fig:fig4} using data and parameters from Italy.
Here, a lockdown is officially ordered on Mar~10 ($\tau^*$, but its late effect on the transmission dynamics starts on Mar~15 ($\hat{\tau}$). If health officials assume the dynamics to immediately change at $\tau^*$, they will expect the number of cases to follow the dashed red line. However, the number of cases will actually follow the black line, leading to a significant different ($\Delta$) between the projections and the realization.

%%% Fig 4 %%%
\begin{figure}[h]
    \centering
	\includegraphics[width=0.5\textwidth]{../figures/Fig4.pdf}
    \caption{
    \textbf{Late effective start of NPIs leads to under-estimation of daily confirmed cases.}
    Real number of daily cases in Italy in black (markers: data, line: time moving average). Model predictions, assuming a 50\% decrease in transmission rate after the NPI starts, are shown as colored lines with 95\% confidence intervals. 
    Shaded box illustrates a serial interval of seven days.
    \textbf{(A)} Using the official date $\tau^*$ for the start of the NPI,  the model under-estimates the number of cases seven days after the start of the NPI.
    \textbf{(B)} Using the effective date $\hat{\tau}$ for the start of the NPI,  the model correctly estimates the number of cases seven days after the start of the NPI.
    Here, model parameters are estimates for Italy (\autoref{table:table2}) but with $\lambda=0.5$ and $\alpha_1=\alpha_2$.
    } 
    \label{fig:fig4}
\end{figure}

@@

\hl{Attempts to asses the effect} of NPIs~\citep{Flaxman2020,Banholzer2020} generally assume a 7 day delay between the implementation of the intervention and the observable change in dynamics, due to the characteristic serial interval of COVID\=/19~\citep{Gatto2020}.
However, the late and early effects we have estimated can confuse these assessments and lead to wrong conclusions about the effects of NPIs (\autoref{fig:fig4}).

>>>>

However, a significant difference between the beginning of the intervention and the effective change in transmission rates can invalidate assessments that assume a serial interval of 4-7 days and neglect the late or early population response to the NPI.
This is illustrated in \autoref{fig:fig4} using data and parameters from Italy.
Here, a lockdown is officially ordered on Mar~10 ($\tau^*$), but its late effect on the transmission dynamics starts on Mar~16 ($\hat{\tau}$). If health officials assume the dynamics to immediately change at $\tau^*$, they will expect the number of cases be within the red lines (posterior predictions assuming $\tau=\tau^*$).
This leads to a significant underestimation, which might be interpreted by officials as ineffectiveness of NPIs, leading to further escalations.
However, the number of cases will actually follow the blue lines (posterior predictions using $\tau=\hat{\tau}$), which corresponds well to the real data.

%%% Fig 4 %%%
\begin{figure}[h]
    \centering
	\includegraphics[width=0.5\textwidth]{../figures/Fig4.pdf}
    \caption{
    \textbf{Late effective start of NPIs leads to under-estimation of daily confirmed cases.}
    Real number of daily cases in Italy in black (markers: data, line: time moving average). 
    Model posterior predictions are shown as colored lines (1,000 draws from the posterior distribution). 
    Shaded box illustrates a serial interval of seven days.
    \textbf{(A)} Using the official date $\tau^*$ for the start of the NPI,  the model under-estimates the number of cases seven days after the start of the NPI.
    \textbf{(B)} Using the effective date $\hat{\tau}$ for the start of the NPI,  the model correctly estimates the number of cases seven days after the start of the NPI.
    Here, model parameters are estimates for Italy (\autoref{table:table2}).
    } 
    \label{fig:fig4}
\end{figure}

@@

Attempts to asses the effect of NPIs~\citep{Flaxman2020,Banholzer2020} generally assume a seven days delay between the implementation of the intervention and the observable change in dynamics, due to the characteristic serial interval of COVID\=/19~\citep{Gatto2020}.
However, the late and early effects we have estimated can confuse these assessments and lead to wrong conclusions about the effects of NPIs (\autoref{fig:fig4}).
